# Scenario X: Chi-Squared Distribution {#scenarioX}

## Details

In this scenario, we consider the case where the test statistic is Chi-squared distributed. It is useful when we conduct a test with more than two treatment arms or when we test two-sided with a normally distributed test statistic. In the first case, we assume binary endpoints, so we have a $2 \times k$ contingency table. In the second case, we conduct a two-sided Z-test. As the normal distribution is symmetric around $0$, this test is equivalent to a one-sided test with test statistic $Z^2$.

## Variant X-1: Contingeny table with binary endpoints

### Setup

Under the alternative, we assume a response rate of $0.4$ in the first group, $0.5$ in the second and $0.6$ in the third group.
```{r}
rate_vec <- c(0.4, 0.5, 0.6)
theta <- get_tau_Pearson2xK(rate_vec)
datadist <- Pearson2xK(3)
H_0 <- PointMassPrior(.0, 1)
prior <- PointMassPrior(theta, 1)

alpha <- 0.025
min_power <- 0.9
toer_cnstr <- Power(datadist, H_0) <= alpha
pow_cnstr <- Power(datadist,prior) >= min_power

```

### Objective

The expected sample size under the alternative shall be minimized.
```{r}
ess <- ExpectedSampleSize(datadist, prior)
```

### Initial design

The initial designs for the minimization process is chosen using the in-built function.

```{r}
tbl_designs <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "one-stage"),
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "group-sequential"),
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "two-stage")))
        
```

### Optimization

```{r}
tbl_designs <- tbl_designs %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr
          ),
          
          initial_design = ., 
          opts           = opts)) )
```


### Test Cases

We first verify that the number of iterations was not exceeded in any of the three cases.

```{r}
tbl_designs %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```

We then check via simulation that the type I error and power constraints are fulfilled.

```{r}
tbl_designs %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], theta , datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power * (1 - tol))) }
```

We expect the sample size function $n_2$ to be monotonously decreasing.
```{r}
testthat::expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```


Since the degrees of freedom of the three design classes are ordered as
'two-stage' > 'group-sequential' > 'one-stage',
the expected sample sizes (under the alternative) should be ordered 
in reverse ('two-stage' smallest).
Additionally, expected sample sizes under both null and alternative
are computed both via `evaluate()` and simulation-based.

```{r}
ess0 <- ExpectedSampleSize(datadist, H_0)

tbl_designs %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, theta, datadist)$n ),
        ess0     = map_dbl(optimal,
                           ~evaluate(ess0, .$design) ),
        ess0_sim = map_dbl(optimal,
                           ~sim_n(.$design, .0, datadist)$n ) ) %>% 
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # sim/evaluate same under null?
    testthat::expect_equal(.$ess0, .$ess0_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```

The expected sample size under the alternative of the optimized designs should be lower than the sample size of the initial designs.

```{r}
testthat::expect_lte(
  evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(optimal) %>% 
                .[[1]]  %>%
                .$design ),
    evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(initial) %>% 
                .[[1]] ) )
```

## Variant X-2: Two-sided Z-test

In the second variant, we want to conduct a two-sided test with normally distributed outcome. Unfortunately, a two-sided test in the setting of two-stage trials is more difficult to implement, as the decision whether to stop early or to continue cannot be expressed in the same way as usual ($Z < c_f$ for futility, $Z > c_e$ for efficacy, $Z \in [c_f, c_e]$ for continuation). In the two-sided test scenario, we would need to consider $|Z|$, so $|Z| < c_f$ for futility, $|Z| > c_e$ for efficacy, $|Z| \in [c_f, c_e]$ for continuation. Unfortunately, this would lead to $4$ necessary boundaries, which would not fit the adoptr-framework. 
If $Z \sim \mathcal{N}(0,1)$ under the null, we can use the fact $Z^2 \sim \chi^2$ and transform our two-sided test problem to a one-sided test. Hence, we stop early for futility if $Z^2 < c_f$, for efficacy if $Z^2 > c_e$ and we continue if $Z^2 \in [c_f, c_e]$, where $c_f$ and $c_e$ are calculated by adoptr using a $\chi^2$-distribution.

### Setup

Under the alternative, we assume an effect size of $0.4$. We want to get a Type I error $\alpha \leq 0.05$ and our power should be higher than $0.8$

```{r}
theta <- get_tau_ZSquared(0.4, 1)
datadist <- ZSquared(two_armed = TRUE)
H_0 <- PointMassPrior(.0, 1)
prior <- PointMassPrior(theta, 1)

alpha <- 0.05
min_power <- 0.8
toer_cnstr <- Power(datadist, H_0) <= alpha
pow_cnstr <- Power(datadist,prior) >= min_power
```

### Objective

We minimize the expected sample size under the alternative.
```{r}
ess <- ExpectedSampleSize(datadist, prior)
```

### Initial designs

The initial designs for the minimization process is chosen using the in-built function.
```{r}
tbl_designs <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "one-stage"),
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "group-sequential"),
        get_initial_design(theta, alpha, 1- min_power, 
                           dist = datadist, type_design = "two-stage")))
```

### Optimization

```{r}
tbl_designs <- tbl_designs %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr
          ),
          
          initial_design = ., 
          opts           = opts)) )
```

### Test

We check that in none of the three cases the number of iterations was exceeded.
```{r}
tbl_designs %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```

We now verify the type I error and power constraints. To get better insights, we do not simulate the test statistic, but normally distributed outcomes and then square the test statistic.
```{r}

N <- 10^6
simulation <- function(design, theta, dist){
    count <- 0
    for(i in 1:N){
        n1 <- n1(design)
        treatment <- rnorm(n1, mean = theta)
        control <- rnorm(n1)
    
        Z_square <- (sqrt(n1 / 2) * (mean(treatment) - mean(control)))^2
    
        if(Z_square > design@c1e){
            count <- count + 1
        }
    
        if(Z_square >= design@c1f & Z_square <= design@c1e){
            n2 <- n2(design, Z_square)
            treatment2 <- rnorm(n2, mean = theta)
            control2 <- rnorm(n2)
    
            Z_final <- (sqrt(n2 / 2) * (mean(treatment2) - mean(control2)))^2
            if(Z_final > c2(design, Z_square)){
                count <- count + 1
            }
        }
    }
    return(list(
        prob = count / N))
}


tbl_designs %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs$optimal, 
                         ~simulation(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs$optimal, 
                         ~simulation(.[[1]], .4, datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power * (1 - tol))) }
```

The $n_2$ function should be monotonously decreasing.

```{r}
testthat::expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```

Since the degrees of freedom of the three design classes are ordered as
'two-stage' > 'group-sequential' > 'one-stage',
the expected sample sizes (under the alternative) should be ordered 
in reverse ('two-stage' smallest).
Additionally, expected sample sizes under both null and alternative
are computed both via `evaluate()` and simulation-based.

```{r}
ess0 <- ExpectedSampleSize(datadist, H_0)

tbl_designs %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, theta, datadist)$n ),
        ess0     = map_dbl(optimal,
                           ~evaluate(ess0, .$design) ),
        ess0_sim = map_dbl(optimal,
                           ~sim_n(.$design, .0, datadist)$n ) ) %>% 
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # sim/evaluate same under null?
    testthat::expect_equal(.$ess0, .$ess0_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```

The optimized design should have a lower expected sample size than the initial designs.
```{r}
testthat::expect_lte(
  evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(optimal) %>% 
                .[[1]]  %>%
                .$design ),
    evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(initial) %>% 
                .[[1]] ) )
```

