# Scenario XI: F-Distribution {#scenarioXI}

## Details

In this scenario, we compute the necessary sample size for a one-way ANOVA with $k$ groups. The resulting test statistic is $F$-distributed with $k-1$ and $nk-k$ degrees of freedom, where $n$ stands for the sample size per group.

## Variant XI-1: ANOVA

### Setup

Under the alternative, we assume a mean of $0.0$ in the first group, $0.4$ in the second group and $0.2$ in the third group.

```{r}
means <- c(0, 0.4, 0.2)
theta <- get_tau_ANOVA(means)

datadist <- ANOVA(3)
H_0 <- PointMassPrior(0, 1)
prior <- PointMassPrior(theta, 1)

alpha <- 0.05
min_power <- 0.8

toer_cnstr <- Power(datadist, H_0)  <= alpha
pow_cnstr <- Power(datadist, prior) >= min_power
```

### Objective

We want to minimize the expected sample size under the alternative.
```{r}
ess <- ExpectedSampleSize(datadist, prior)
```

### Initial Design

The initial designs are chosen via the in-built function to create initial designs.

```{r}
tbl_designs <- tibble(
    type    = c("one-stage", "group-sequential", "two-stage"),
    initial = list(
        get_initial_design(theta, alpha, 1 - min_power, 
                           "one-stage", dist = datadist),
        get_initial_design(theta, alpha, 1 - min_power, 
                           "group-sequential", dist = datadist),
        get_initial_design(theta, alpha, 1 - min_power, 
                           "two-stage", dist = datadist) ))
```

### Optimization

```{r}
tbl_designs <- tbl_designs %>% 
    mutate(
       optimal = purrr::map(initial, ~minimize(
         
          ess,
          subject_to(
              toer_cnstr,
              pow_cnstr 
          ),
          
          initial_design = ., 
          opts           = opts)) )
```

### Test Cases

We first verify that in none of the three cases, the maximal number of iterations was exceeded.
```{r}
tbl_designs %>% 
  transmute(
      type, 
      iterations = purrr::map_int(tbl_designs$optimal, 
                                  ~.$nloptr_return$iterations) ) %>%
  {print(.); .} %>% 
  {testthat::expect_true(all(.$iterations < opts$maxeval))}
```

We then check via simulations of the test statistic, that the type I error and power constraints are fulfilled.

```{r}
tbl_designs %>% 
  transmute(
      type, 
      toer  = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], .0, datadist)$prob), 
      power = purrr::map(tbl_designs$optimal, 
                         ~sim_pr_reject(.[[1]], theta , datadist)$prob) ) %>% 
  unnest(., cols = c(toer, power)) %>% 
  {print(.); .} %>% {
  testthat::expect_true(all(.$toer  <= alpha * (1 + tol)))
  testthat::expect_true(all(.$power >= min_power * (1 - tol))) }
```

The sample size function $n_2$ should be monotonously decreasing.

```{r}
testthat::expect_true(
    all(diff(
        # get optimal two-stage design n2 pivots
        tbl_designs %>% filter(type == "two-stage") %>%
           {.[["optimal"]][[1]]$design@n2_pivots} 
        ) < 0) )
```

Since the degrees of freedom of the three design classes are ordered as
'two-stage' > 'group-sequential' > 'one-stage',
the expected sample sizes (under the alternative) should be ordered 
in reverse ('two-stage' smallest).
Additionally, expected sample sizes under both null and alternative
are computed both via `evaluate()` and simulation-based.

```{r}
ess0 <- ExpectedSampleSize(datadist, H_0)

tbl_designs %>% 
    mutate(
        ess      = map_dbl(optimal,
                           ~evaluate(ess, .$design) ),
        ess_sim  = map_dbl(optimal,
                           ~sim_n(.$design, theta, datadist)$n ),
        ess0     = map_dbl(optimal,
                           ~evaluate(ess0, .$design) ),
        ess0_sim = map_dbl(optimal,
                           ~sim_n(.$design, .0, datadist)$n ) ) %>% 
    {print(.); .} %>% {
    # sim/evaluate same under alternative?
    testthat::expect_equal(.$ess, .$ess_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # sim/evaluate same under null?
    testthat::expect_equal(.$ess0, .$ess0_sim, 
                           tolerance = tol_n,
                           scale = 1)
    # monotonicity with respect to degrees of freedom
    testthat::expect_true(all(diff(.$ess) < 0)) }
```

The expected sample size under the alternative of the optimized designs should be lower than the sample size of the initial designs.

```{r}
testthat::expect_lte(
  evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(optimal) %>% 
                .[[1]]  %>%
                .$design ),
    evaluate(ess, 
             tbl_designs %>% 
                dplyr::pull(initial) %>% 
                .[[1]] ) )
```


